---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# survatr

<!-- badges: start -->
[![R-CMD-check](https://github.com/etverse/survatr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/etverse/survatr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/etverse/survatr/graph/badge.svg)](https://app.codecov.io/gh/etverse/survatr)
<!-- badges: end -->

**survatr** extends the [causatr](https://github.com/etverse/causatr) engine to
**time-to-event outcomes**. It estimates counterfactual survival via
pooled-logistic discrete-time hazard models on person-period data, with three
complementary estimators: g-computation (Track A), inverse-probability weighting
(IPW, Track A), and iterated conditional expectations (ICE) for longitudinal,
time-varying treatments (Track B). The **estimand is a curve** — survival
$S^a(t)$, risk $1 - S^a(t)$, risk difference, risk ratio, restricted mean
survival time (RMST) or time lost (RMTL), survival quantiles / median, and
per-cause years of life lost — not a scalar. Inference is by a delta-method
sandwich
that propagates per-row influence functions through the cumulative-product
survival curve, or by an individual-level nonparametric bootstrap.

The package implements the survival methods of Hernán & Robins (2025) *Causal
Inference: What If* (Chapter 17) with the same two-step API causatr users know:

1.  **Fit** the hazard model once with `surv_fit()`
2.  **Contrast** many interventions cheaply with `contrast()`

## Installation

You can install the development version of survatr from
[GitHub](https://github.com/etverse/survatr) with:

``` r
# install.packages("pak")
pak::pak("etverse/survatr")
```

## Quick example

The data-generating process below has a constant 6% discrete-time hazard, a
single binary treatment with no causal effect, and one confounder `L` that
drives both treatment and the hazard. Under this DGP the counterfactual survival
under either intervention converges to $(1 - 0.06)^t$.

```{r example, message = FALSE}
library(causatr) # attach first so survatr's contrast() generic shadows it
library(survatr)
library(data.table)

# Simulate person-period data: constant hazard, one confounder L.
set.seed(1)
n <- 800L
K <- 10L
L <- rnorm(n)
A <- rbinom(n, 1L, plogis(0.3 * L)) # L affects treatment
pp <- rbindlist(lapply(seq_len(n), function(i) {
  Y <- rbinom(K, 1L, 0.06) # hazard independent of A (null effect)
  first <- which(Y == 1L)[1L]
  if (!is.na(first) && first < K) Y[(first + 1L):K] <- 0L
  data.table(id = i, t = seq_len(K), A = A[i], L = L[i], Y = Y)
}))

# Step 1: fit the pooled-logistic hazard model once.
fit <- surv_fit(
  pp,
  outcome      = "Y",
  treatment    = "A",
  confounders  = ~ L,
  id           = "id",
  time         = "t",
  time_formula = ~ factor(t) # non-parametric baseline hazard
)

# Step 2: contrast interventions on the fitted hazard. Risk difference with
# a delta-method sandwich CI at each of the 10 periods.
res <- contrast(
  fit,
  interventions = list(a1 = static(1), a0 = static(0)),
  times         = 1:10,
  type          = "risk_difference",
  reference     = "a0",
  ci_method     = "sandwich"
)
res$contrasts[] # `[]` forces data.table to auto-print a function-returned table
```

The DGP has no treatment effect, so the risk-difference CI comfortably covers 0
at every time point.

## Features

- **Three estimators**: g-computation (`estimator = "gcomp"`, pooled-logistic
  standardization), IPW (`estimator = "ipw"`, a weighted marginal hazard MSM
  with stabilized density-ratio weights), and ICE (`estimator = "ice"`, Track B
  longitudinal survival via backward iterated conditional expectations for a
  time-varying treatment). Matching + survival is a hard reject — use
  `survival::coxph(..., weights, cluster)` directly.
- **Curve-valued estimands**: survival $S^a(t)$, risk $1 - S^a(t)$, risk
  difference, risk ratio (log-scale CI), RMST / RMST difference, RMTL
  (restricted mean time lost), survival quantiles / median (implicit-function
  delta, `q`-indexed), and per-cause years of life lost (competing risks), all
  returned as `data.table`s.
- **Flexible interventions** via causatr: `static()`, `shift()`, `scale_by()`,
  `threshold()`, and `dynamic()`. Which are available depends on the estimator.
- **Robust inference**: a delta-method sandwich that aggregates per-individual
  influence functions across time through the cumulative product (pointwise Wald
  bands; stacked estimating equations for IPW and ICE), or a nonparametric
  bootstrap that resamples ids and refits per replicate.
- **Pluggable hazard models** via `model_fn` (`stats::glm` by default;
  `mgcv::gam` with an `s(time)` baseline-hazard term, splines via `ns()`).
- **`tidy()` / `plot()` / `print()` / `forrest()`** S3 methods that dispatch on
  the time-indexed `survatr_result` shape (survival curves, and forest plots at
  a reference time `t*`).

## Learning more

- `vignette("survatr")` — introduction and the two-step API
- `vignette("ipw")` — Track A IPW (weighted hazard MSM)
- `vignette("ice")` — Track B longitudinal survival via ICE

## References

Hernán MA, Robins JM (2025). *Causal Inference: What If*. Chapman & Hall/CRC.
Chapter 17 (survival analysis via IP weighting and standardization).

Zivich PN, Ross RK, Shook-Sa BE, Cole SR, Edwards JK (2024). Empirical sandwich
variance estimator for iterated conditional expectation g-computation.
*Statistics in Medicine* 43:5562–5572.

## Acknowledgements

This package was built with the contribution of [Claude](https://claude.ai),
Anthropic's AI assistant.
