---
title: "Causal survival analysis: NHEFS Ch. 17 walkthrough"
code-fold: show
code-tools: true
vignette: >
  %\VignetteIndexEntry{Causal survival analysis: NHEFS Ch. 17 walkthrough}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
## Skip heavy computations on CRAN (NOT_CRAN not set).
NOT_CRAN <- identical(Sys.getenv("NOT_CRAN"), "true")
knitr::opts_chunk$set(eval = NOT_CRAN)
```

This vignette walks through the Hernán & Robins *Causal Inference: What If*
Chapter 17 analysis using survatr's two-step API. The analysis estimates the
effect of smoking cessation (`qsmk`) on 10-year all-cause mortality in the
NHEFS cohort.

**Reference target** (H&R 2024, Program 17.2): 120-month standardised survival
≈ 80.7% under cessation, ≈ 80.5% under no cessation; risk difference ≈ 0.2%
(95% CI −4.1% to 3.7% — essentially null).

## Data

The `nhefs_surv` dataset ships with survatr. It was assembled from
`causaldata::nhefs` (Huntington-Klein 2021) by expanding each of the 1629
participants to a rectangular person-period grid of 120 monthly rows.

```{r}
## Load causatr first so survatr::contrast() shadows causatr's scalar-outcome
## contrast() and dispatches correctly on survatr_fit objects.
library(causatr)
library(survatr)
library(data.table)

data(nhefs_surv)
nhefs_surv
```

Each row is one *person-period*: one individual in one calendar month. The
`event` column is 1 only in the month of death for the 318 participants who died
before December 1992; it is 0 in all other rows. The nine confounders from
Table 17.1 are carried as baseline values across all rows for each participant.

## Step 1 — Fit the hazard model

The discrete-time hazard model follows Program 17.2: a pooled logistic
regression of `event` on `qsmk`, the nine confounders, a quadratic time
baseline, and treatment-by-time interactions.

```{r}
#| cache: true
fit <- surv_fit(
  nhefs_surv,
  outcome      = "event",
  treatment    = "qsmk",
  confounders  = ~ sex + race + age + I(age^2) +
    education +
    smokeintensity + I(smokeintensity^2) +
    smokeyrs + I(smokeyrs^2) +
    exercise + active + wt71 + I(wt71^2),
  id           = "seqn",
  time         = "time",
  ## Quadratic time baseline + treatment-by-time interactions (H&R Ch. 17).
  time_formula = ~ time + I(time^2) + qsmk:time + qsmk:I(time^2),
  estimator    = "gcomp"
)
fit
```

## Step 2 — Survival curves and risk difference

`contrast()` takes the fitted model and returns time-indexed estimates and
contrasts under each intervention.

```{r}
#| cache: true
times <- seq(12L, 120L, by = 12L)

res <- contrast(
  fit,
  interventions = list(
    cessation    = static(1),
    no_cessation = static(0)
  ),
  times     = times,
  type      = "survival",
  ci_method = "sandwich"
)

## Standardised survival curves at each annual checkpoint.
res$estimates
```

The risk difference under cessation vs. no cessation at 120 months:

```{r}
#| cache: true
rd <- contrast(
  fit,
  interventions = list(
    cessation    = static(1),
    no_cessation = static(0)
  ),
  times     = 120L,
  type      = "risk_difference",
  ci_method = "sandwich"
)
rd$contrasts
```

The point estimate (≈ 0.2 pp) and the sandwich 95% CI both agree with the
published analysis: smoking cessation has essentially no detectable effect on
10-year mortality in this observational cohort after confounder adjustment.

## RMST

The restricted mean survival time (RMST) up to 120 months summarises the
survival curve as a single number: the area under $\hat{S}^a(t)$ from 0 to
$t^* = 120$.

```{r}
#| cache: true
rmst <- contrast(
  fit,
  interventions = list(
    cessation    = static(1),
    no_cessation = static(0)
  ),
  times     = times,
  type      = "rmst_difference",
  ci_method = "sandwich"
)
rmst$contrasts
```

## RMTL (restricted mean time lost)

Restricted mean time *lost* is the complement of RMST within the same window,
$\mathrm{RMTL}^a(t^*) = t^* - \mathrm{RMST}^a(t^*) = \int_0^{t^*} (1 - \hat{S}^a(u))\,du$.
It is the same curve viewed from the other side, so the standard error is
identical to RMST's (the constant $t^*$ drops out of the gradient).

```{r}
#| cache: true
rmtl <- contrast(
  fit,
  interventions = list(
    cessation    = static(1),
    no_cessation = static(0)
  ),
  times     = times,
  type      = "rmtl_difference",
  ci_method = "sandwich"
)
rmtl$contrasts
```

## Survival quantiles

The $q$-quantile of survival time is the first $t$ at which $\hat{S}^a(t)$
drops to $1 - q$ (the median is $q = 0.5$). On this NHEFS cohort survival stays
above $\approx 0.8$ through 120 months, so the median is *beyond the follow-up
window* and `q = 0.5` would abort with `survatr_quantile_unreached` — a long
survival horizon is exactly when you report a low quantile (or RMST) instead.
Here we report the 10th and 15th percentiles of survival time; pass a vector of
`q` to get several at once. The result is indexed by `q`, and a difference in
quantile survival time is reported when there are two interventions.

```{r}
#| cache: true
quant <- contrast(
  fit,
  interventions = list(
    cessation    = static(1),
    no_cessation = static(0)
  ),
  times     = times,
  type      = "quantile",
  q         = c(0.10, 0.15),
  ci_method = "sandwich"
)
quant$estimates
```

## Plot

```{r}
#| fig-alt: "Standardised survival curves under smoking cessation and no cessation."
plot(res)
```

## References

- Hernán MA, Robins JM (2024). *Causal Inference: What If.* Boca Raton: Chapman
  & Hall / CRC. Chapter 17: Causal survival analysis.
- Huntington-Klein N et al. (2021). causaldata: Example datasets for causal
  inference textbooks. *R package version 0.1.*
- D'Agostino RB et al. (1990). Relation of pooled logistic regression to
  time dependent Cox regression analysis: the Framingham Heart Study. *Statistics
  in Medicine* 9:1501–1515.
